---
title: "Examining the Relationship Between Air Quality, Poverty, and Suicide Mortality in Michigan and California in 2022."
author: "Isabella Bommarito"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 pdf_document:
     pandoc_args: ["--extract-media", "."]
---
```{r setup, include=TRUE, echo = FALSE}
# Leave this chunk as it is. It sets defaults for subsequent chunks.
# The "echo" option tells Rmarkdown whether or not it should show the code in the chunk.
# Setting echo = TRUE outputs both the code and the results
knitr::opts_chunk$set(echo = TRUE,
                     collapse = TRUE,
                     warning = FALSE, message = FALSE, error = FALSE,
                     fig.align = 'center')
```
**Note: Generative AI was used in this project**

## 1. **Introduction:** 
This project examines how air quality and poverty interact to influence suicide mortality at the county level in California and Michigan. Using merged data from the EPA, CDC, and U.S. Census, we explore how environmental and socioeconomic stressors jointly shape mental health outcomes. We find that counties with both high poverty and frequent PM2.5 pollution days exhibit substantially higher suicide rates. Building on these findings, we estimate a predictive model using LASSO regression to assess whether county-level suicide risk can be forecasted using publicly available data.

## 2. **Data:**
This project draws from four publicly available datasets to examine the relationship between air quality, poverty, and suicide mortality across counties in California and Michigan for the year 2022. The four datasets used include:

Air Quality Data from the EPA, providing daily pollutant measures such as PM2.5, ozone, and AQI indicators.

Mortality Data from the CDC, offering provisional suicide death counts at the county level.

Poverty Estimates from the U.S. Census Bureau, reporting the percentage of the population living below the poverty line.

Population Estimates from the Census Bureau, supplying 2022 county-level population data used to compute suicide rates.

To prepare the data for analysis, we standardized geographic labels and harmonized state names using two-letter abbreviations (e.g., CA for California, MI for Michigan). County names were cleaned to remove suffixes like “County” to allow proper alignment. All four datasets were then merged using a composite key consisting of the county name and state abbreviation. We did not use FIPS codes due to inconsistent availability across sources. We applied an inner_join() strategy to ensure only counties with complete records across all datasets were retained. This merging approach resulted in a unified dataset with consistent geographic granularity.

Handling missing data was a critical part of the cleaning process. We filtered out incomplete or non-numeric entries in the poverty dataset and excluded rows with missing air quality or mortality values. While we aimed to retain as many counties as possible, we prioritized data integrity over volume, ensuring that every included observation had full information across all indicators.

The final dataset includes 54 counties in total. For each county, we calculated the suicide rate per 100,000 residents, rather than relying on raw death counts, to normalize across populations of varying sizes. On average, counties had approximately 164 days exceeding PM2.5 pollution thresholds, a 13% poverty rate, and a suicide rate of 14.81 per 100,000 population.

```{r, echo=FALSE}
library(tidyverse)
library(gssr)
library(readxl)
library(dplyr)
library(glmnet)
library(ggplot2)
library(tidyr)
library(tibble)
library(stringr)

# Step 1: Read in the dataset
air_quality_county <- read_csv("2022_County_AQ.csv")

# Step 2: Filter for Michigan and California
air_quality_county <- air_quality_county %>%
  filter(State %in% c("Michigan", "California"))

# Step 3: Create a new 'Location' column to match mortality dataset format
air_quality_county <- air_quality_county %>%
  mutate(Location = paste(County, "County,", substr(State, 1, 2)))

air_quality_county <- air_quality_county%>%
  select(-State)


#CDC Suicide data

# Step 1: Read in the mortality dataset
mortality_df <- read_excel("Provisional Mortality Statistics_2022_CDC.xlsx", 
                           sheet = "Provisional Mortality Statistic") %>%
  select(`Occurrence County`, `Occurrence County Code`, Deaths) %>%
  filter(!is.na(`Occurrence County`)) %>%
  filter(str_detect(`Occurrence County`, "MI|CA"))

# Step 2: Extract State and clean County name
mortality_df <- mortality_df %>%
  mutate(
    State = str_extract(`Occurrence County`, "[A-Z]{2}$"),                  # Get last 2 capital letters (state)
    County = str_replace(`Occurrence County`, " County, [A-Z]{2}$", "")     # Remove ' County, XX'
  )

# Step 3: Reorder columns
mortality_df_clean <- mortality_df %>%
  select(County, State, Deaths, `Occurrence County Code`)


#Poverty Census data

# Step 1: Read the poverty data (skip header info row)
poverty_data <- read_csv("Poverty_Data_Cencus.csv", skip = 1)

# Find and rename the correct column manually
poverty_data <- poverty_data %>%
  rename(
    Poverty_Raw = `Estimate!!Percent below poverty level!!Population for whom poverty status is determined`
  )

# Continue as before
poverty_data_clean <- poverty_data %>%
  filter(str_detect(`Geographic Area Name`, "County")) %>%
  mutate(
    County = str_remove(`Geographic Area Name`, " County,.*"),
    State = str_extract(`Geographic Area Name`, ",\\s*(\\w+)$") %>% str_remove(",\\s*"),
    Poverty_Rate = as.numeric(Poverty_Raw)
  ) %>%
  select(County, State, Poverty_Rate) %>%
  filter(!is.na(Poverty_Rate)) %>%
  filter(State %in% c("California", "Michigan")) %>%
  mutate(State = recode(State, "California" = "CA", "Michigan" = "MI"))

#Population Data
pop_data <- read_csv("county_population.csv")

# Extract just what you need
# This depends on the exact column names in the file, but often:
pop_data_clean <- pop_data %>%
  filter(STNAME %in% c("California", "Michigan")) %>%
  mutate(
    State = case_when(
      STNAME == "California" ~ "CA",
      STNAME == "Michigan" ~ "MI"
    ),
    County = str_replace(CTYNAME, " County", "")
  ) %>%
  select(County, State, Population = POPESTIMATE2022)



#Merging data

# Step 1: Merge air quality and mortality data (first two datasets)
merged_air_mortality <- inner_join(air_quality_county, mortality_df_clean, by = c("County"))

# Step 2: Add in poverty data (third dataset)
final_merged_data <- inner_join(merged_air_mortality, poverty_data_clean, by = c("County", "State"))

final_merged_data <- inner_join(final_merged_data, pop_data_clean, by = c("County", "State"))
final_merged_data <- final_merged_data %>%
  mutate(Suicide_Rate = (Deaths / Population) * 100000)

```


## 3.**Analysis with Visuals**
```{r, echo=FALSE}

#Poverty Vs. Air Pollution
ggplot(final_merged_data, aes(x = Poverty_Rate, y = `Days PM2.5`, color = State)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("CA" = "#D55E00", "MI" = "#0072B2")) +
  labs(
    title = "Counties with Higher Poverty See More PM2.5 Days",
    x = "Poverty Rate (%)",
    y = "Days with PM2.5 Pollution",
    color = "State"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))

#PM2.5 and Poverty on Suicide by State
final_merged_data <- final_merged_data %>%
  mutate(State = as.factor(State))

final_merged_data <- final_merged_data %>%
  mutate(Poverty_Group = ifelse(Poverty_Rate > median(Poverty_Rate, na.rm = TRUE), "High Poverty", "Low Poverty"))

final_merged_data <- final_merged_data %>%
  mutate(Pollution_Group = ifelse(`Days PM2.5` > median(`Days PM2.5`, na.rm = TRUE), "High Pollution", "Low Pollution"))

# Create combined group
final_merged_data <- final_merged_data %>%
  mutate(Group = paste(Poverty_Group, Pollution_Group, sep = " + "))

ggplot(final_merged_data, aes(x = `Days PM2.5`, y = Suicide_Rate)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "#D55E00") +
  facet_grid(Poverty_Group ~ State) +
  labs(
    title = "Interaction of Pollution and Poverty by State",
    x = "Days with PM2.5 Pollution",
    y = "Suicide Rate (per 100,000)"
  ) +
  theme_minimal(base_size = 14)


#Suicide Deaths by Poverty Group and Pollution Group
# Define Pollution_Group and combined Group
final_merged_data <- final_merged_data %>%
  mutate(Pollution_Group = ifelse(`Days PM2.5` > median(`Days PM2.5`, na.rm = TRUE), 
                                  "High Pollution", "Low Pollution"),
         Group = paste(Poverty_Group, Pollution_Group, sep = " + "))

ggplot(final_merged_data, aes(x = Group, y = Suicide_Rate, fill = Group)) +
  geom_boxplot(width = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, color = "black") +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73", "#D55E00")) +
  labs(
    title = "Suicide Rates by Combined Poverty and Pollution Group",
    x = "Poverty and Pollution Group",
    y = "Suicide Rate (per 100,000)"
  ) +
  theme_minimal(base_size = 8) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 19, hjust = 1),
    legend.position = "none"
  )

```
While multiple air pollutants were available in the dataset (including Ozone, NO2, and CO), we chose to focus our visual analysis on PM2.5 due to its strong documented associations with both physical and mental health outcomes. PM2.5 is widely regarded as one of the most harmful pollutants because it can penetrate deep into the lungs and bloodstream, potentially influencing neurological and psychological systems. Additionally, it showed the most variation across counties in our dataset, making it a meaningful variable for exploring geographic disparities. Though other pollutants were included in the predictive model, PM2.5 served as the primary air quality indicator in our visual exploration to maintain clarity and interpretability.

Socioeconomic disadvantage and air quality appear to interact in shaping suicide mortality patterns across counties. Counties with higher poverty rates tend to experience more days with PM2.5 pollution, indicating overlapping environmental and economic stressors. A positive relationship between PM2.5 days and suicide deaths is particularly evident in high-poverty counties, suggesting that poor air quality may have a more pronounced impact on mental health in disadvantaged areas.

Our first visualization plots Poverty Rate vs. Days Exceeding PM2.5 Standards, with points color-coded by state. A clear positive trend emerges: counties with higher poverty rates tend to experience more days of unsafe PM2.5 levels. This relationship appears in both California and Michigan, suggesting that economically disadvantaged areas face disproportionate exposure to harmful air pollution — consistent with known patterns of environmental inequality.

The second figure explores how PM2.5 exposure correlates with suicide rates, split by state and poverty level (above/below the median). The most striking trend appears in high-poverty counties, where a strong positive relationship is observed: suicide rates rise sharply with increased PM2.5 exposure. In contrast, this relationship is weaker or absent in low-poverty counties. This divergence suggests a conditional effect, where the mental health burden of air pollution is amplified by economic vulnerability.

The third visualization compares suicide rates across four combined groups: High/Low Poverty × High/Low PM2.5. The “High Poverty + High Pollution” group displays the highest median suicide rate and greatest variability, while the “Low Poverty + Low Pollution” group exhibits the lowest and most stable rates. Interestingly, the “High Poverty + Low Pollution” group shows moderately elevated suicide rates, though lower than their high-pollution counterparts. This suggests that poverty alone carries substantial suicide risk, but pollution may act as an additional compounding factor. The gap between high-poverty counties with low vs. high pollution reinforces the idea that environmental exposures intensify existing socioeconomic vulnerabilities.

Together, these visuals reveal a coherent pattern: the intersection of poverty and poor air quality significantly elevates suicide risk. The consistency of this effect across multiple visual formats — including scatterplots, faceted trends, and groupwise comparisons — supports a robust interpretation. These findings contribute to the broader understanding of environmental injustice and suggest that suicide prevention strategies must consider both social and environmental determinants of mental health.


## 4.**Predictive Model**

```{r, echo=FALSE}

# Select predictors and outcome
model_data <- final_merged_data %>%
  select(Suicide_Rate, `Max AQI`, `Days PM2.5`, `Days Ozone`, `Days NO2`, `Days CO`,
         `Median AQI`, Poverty_Rate, State) %>%
  drop_na()  # Remove missing rows

# Convert categorical variables
model_data$State <- as.factor(model_data$State)
model_data$log_Suicide_Rate <- log1p(model_data$Suicide_Rate)


set.seed(0419)  # for reproducibility
n <- nrow(model_data)
eval_index <- sample(1:n, size = floor(0.1 * n))

train_data <- model_data[-eval_index, ]
test_data  <- model_data[eval_index, ]

# Include all pairwise interactions
x_train <- model.matrix(Suicide_Rate ~ .^2, data = train_data)[, -1]
y_train <- log1p(train_data$Suicide_Rate)

x_test <- model.matrix(Suicide_Rate ~ .^2, data = test_data)[, -1]
y_test <- log1p(test_data$Suicide_Rate)

set.seed(0419)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1)
# Optimal lambda
best_lambda <- cv_lasso$lambda.min

train_rmse <- sqrt(cv_lasso$cvm)
lambdas <- cv_lasso$lambda

rmse_df <- data.frame(
  Lambda = lambdas,
  RMSE = train_rmse
)

# Plot RMSE vs. log(lambda)
ggplot(rmse_df, aes(x = log(Lambda), y = RMSE)) +
  geom_line(color = "#0072B2", linewidth = 1.2) +
  geom_vline(xintercept = log(best_lambda), linetype = "dashed", color = "#D55E00") +
  labs(
    title = "Cross-Validated RMSE Across Lambda Values",
    x = "log(Lambda)",
    y = "RMSE"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))

# Predict on test set
lasso_best <- glmnet(x_train, y_train, alpha = 1, lambda = best_lambda)
pred_test_log <- predict(lasso_best, s = best_lambda, newx = x_test)
pred_test <- expm1(pred_test_log)

# Final RMSE on evaluation set
eval_rmse <- sqrt(mean((y_test - pred_test)^2))
print(paste("Evaluation RMSE:", round(eval_rmse, 2)))


actual_test <- expm1(y_test)  # back-transform to original scale

plot(y_test, pred_test, main = "Predicted vs. Actual Suicide Rate",
     xlab = "Actual Suicide Rate", ylab = "Predicted Suicide Rate",
     pch = 19, col = "#0072B2")
abline(0, 1, col = "red", lwd = 2)



```

To predict suicide rates at the county level, we estimated a LASSO regression model using environmental and socioeconomic indicators. The model included daily pollutant measures (PM2.5, ozone, CO, NO2), maximum and median AQI values, poverty rate, and state, along with all possible pairwise interactions to account for complex interdependencies among variables. The outcome variable—suicide rate per 100,000 residents—was log-transformed using log1p() to stabilize variance and reduce right-skewness. We randomly withheld 10% of counties as an evaluation sample and applied 10-fold cross-validation on the training set to select the optimal regularization parameter. The cv.glmnet() function identified a lambda value of 0.009491, which minimized cross-validated RMSE across a log-scaled sequence of penalty values. When applied to the evaluation sample, the model achieved an RMSE of 16.32, meaning predicted suicide rates deviated from actual rates by approximately 16.3 deaths per 100,000 residents on average. Compared to the dataset’s mean suicide rate of 14.81, this error reflects a moderately high deviation, highlighting the difficulty of predicting rare but important public health outcomes with limited regional-level data. As expected, the RMSE curve followed a U-shaped trend, where both very low and very high values of lambda produced higher error, with the optimal point balancing model complexity and regularization.


## 5. **Conclusion**
This project examined how environmental and socioeconomic conditions jointly influence suicide mortality across 54 counties in California and Michigan. By integrating public datasets on air quality, poverty, and suicide deaths, we uncovered consistent patterns linking higher poverty and more frequent PM2.5 pollution days to elevated suicide rates.

Visual analysis revealed a strong association between poverty and pollution exposure, and a compounding effect on suicide risk when both stressors were present. Counties experiencing both high poverty and high pollution exhibited the highest median suicide rates, while low-poverty, low-pollution counties saw the lowest and most stable rates. This points to a powerful interaction effect—poverty and environmental stressors do not act independently, but rather reinforce one another in shaping mental health outcomes.

Our predictive model, built using LASSO regression with interaction terms and a log-transformed outcome, achieved an evaluation RMSE of 16.32—just above the average suicide rate of 14.81 per 100,000. While this suggests a moderate prediction error, the model still captured meaningful structure in the data using publicly available county-level features.

Overall, these results support a broader understanding of environmental injustice, showing that disadvantaged communities are not only more exposed to harmful pollutants but also more vulnerable to their consequences. Suicide prevention efforts must account for these structural conditions, recognizing that improving environmental and economic well-being may also reduce mental health disparities.

